### 2025-07-10 factor-analysis.qmd
20250710 - factor analysis

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_weekly.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_nextGenStats_weekly.RData", fsep = ""))
```

```{r}
load(file = "./data/player_stats_weekly.RData")
load(file = "./data/nfl_nextGenStats_weekly.RData")
```

@@

@@

```{r}
dataMerged <- full_join(
  player_stats_weekly,
  nfl_nextGenStats_weekly %>% select(-any_of(c("player_display_name","completions","attempts","receptions","targets"))),
  by = c("player_id" = "player_gsis_id","season","season_type","week"),
)
```

@@

```{r}
faVars <- c(
  "completions","attempts","passing_yards","passing_tds","passing_interceptions",
  "sacks_suffered","sack_yards_lost","sack_fumbles","sack_fumbles_lost",
  "passing_air_yards","passing_yards_after_catch","passing_first_downs",
  "passing_epa","passing_cpoe","passing_2pt_conversions","pacr","pass_40_yds",
  "pass_inc","pass_comp_pct","fumbles","two_pts",
  "avg_time_to_throw","avg_completed_air_yards","avg_intended_air_yards",
  "avg_air_yards_differential","aggressiveness","max_completed_air_distance",
  "avg_air_yards_to_sticks","passer_rating", #,"completion_percentage"
  "expected_completion_percentage","completion_percentage_above_expectation",
  "avg_air_distance","max_air_distance")
```

@@

Although standardizing variables is not strictly necessary in factor analysis, it can be helpful to prevent some variables from having considerably larger variances than others.

```{r}
dataForFA <- dataMerged
dataForFA[faVars] <- scale(dataForFA[faVars])
```

@@

@@

screeDataEFA <- nFactors::nScree(
  x = cor(
    dataForFA[faVars],
    use = "pairwise.complete.obs"),
  model = "factors")

@@

@@

efaOblique_fit <- lavaan::efa(
  data = dataForFA,
  ov.names = faVars,
  nfactors = 1:9,
  rotation = "geomin",
  missing = "ML",
  estimator = "MLR",
  cluster = "player_id",
  meanstructure = TRUE
)
```

A path diagram of the EFA model was created using the `plot_lavaan()` function of the `lavaangui` package [@R-lavaangui; @Karch2025_packages].

@@

lavaangui::plot_lavaan(efaOblique_fit)
```

@@

@@

```{r}
cfaModel1_syntax <- '
 #Factor loadings
 F1 =~ completions + attempts + passing_yards + passing_tds + passing_interceptions + sacks_suffered + sack_yards_lost + sack_fumbles + sack_fumbles_lost + passing_air_yards + passing_yards_after_catch + passing_first_downs + pass_inc + fumbles
 F2 =~ passing_air_yards + pacr + avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + avg_air_yards_to_sticks + expected_completion_percentage + avg_air_distance + max_air_distance
 F3 =~ passing_tds + passing_epa + passing_cpoe + pass_40_yds + pass_comp_pct + avg_air_yards_differential + max_completed_air_distance + avg_time_to_throw + avg_completed_air_yards + avg_intended_air_yards + passer_rating + completion_percentage_above_expectation
'
```

@@

cfaModel1_fit <- lavaan::cfa(
  cfaModel1_syntax,
  data = dataForFA,
  missing = "ML",
  estimator = "MLR",
  std.lv = TRUE,
  cluster = "player_id",
  em.iter.max = 2000000)
```

@@

summary(
  cfaModel1_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

@@

lavaan::fitMeasures(
  cfaModel1_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

@@

lavaan::residuals(
  cfaModel1_fit,
  type = "cor")
```

lavaan::modificationindices(
  cfaModel1_fit,
  sort. = TRUE)
```

lavaan::lavPredict(cfaModel1_fit)
```

A path diagram of the CFA model was created using the `plot_lavaan()` function of the `lavaangui` package [@R-lavaangui; @Karch2025_packages].

lavaangui::plot_lavaan(cfaModel1_fit)
```

>>>>

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_seasonal.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_weekly.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_nextGenStats_weekly.RData", fsep = ""))
```

```{r}
load(file = "./data/player_stats_seasonal.RData")
load(file = "./data/player_stats_weekly.RData")
load(file = "./data/nfl_nextGenStats_weekly.RData")
```

@@

#### Season-Averages {#sec-factorAnalysisPrepareDataSeasonAvg}

```{r}
player_stats_seasonal_avgPerGame <- player_stats_seasonal %>% 
  mutate(
    completionsPerGame = completions / games,
    attemptsPerGame = attempts / games,
    passing_yardsPerGame = passing_yards / games,
    passing_tdsPerGame = passing_tds / games,
    passing_interceptionsPerGame = passing_interceptions / games,
    sacks_sufferedPerGame = sacks_suffered / games,
    sack_yards_lostPerGame = sack_yards_lost / games,
    sack_fumblesPerGame = sack_fumbles / games,
    sack_fumbles_lostPerGame = sack_fumbles_lost / games,
    passing_air_yardsPerGame = passing_air_yards / games,
    passing_yards_after_catchPerGame = passing_yards_after_catch / games,
    passing_first_downsPerGame = passing_first_downs / games,
    #passing_epaPerGame = passing_epa / games,
    #passing_cpoePerGame = passing_cpoe / games,
    passing_2pt_conversionsPerGame = passing_2pt_conversions / games,
    #pacrPerGame = pacr / games
    pass_40_ydsPerGame = pass_40_yds / games,
    pass_incPerGame = pass_inc / games,
    #pass_comp_pctPerGame = pass_comp_pct / games,
    fumblesPerGame = fumbles / games
  )

nfl_nextGenStats_seasonal <- nfl_nextGenStats_weekly %>% 
  filter(season_type == "REG") %>% 
  select(-any_of(c("week","season_type","player_display_name","player_position","team_abbr","player_first_name","player_last_name","player_jersey_number","player_short_name"))) %>% 
  group_by(player_gsis_id, season) %>% 
  summarise(
    across(everything(),
      ~ mean(.x, na.rm = TRUE)),
    .groups = "drop")
```

@@

```{r}
dataMerged <- full_join(
  player_stats_seasonal_avgPerGame,
  nfl_nextGenStats_seasonal %>% select(-any_of(c("player_display_name","completions","attempts","receptions","targets"))),
  by = c("player_id" = "player_gsis_id","season"),
)
```

@@

c(
  "completionsPerGame","attemptsPerGame","passing_yardsPerGame","passing_tdsPerGame","passing_interceptionsPerGame",
  "sacks_sufferedPerGame","sack_yards_lostPerGame","sack_fumblesPerGame","sack_fumbles_lostPerGame",
  "passing_air_yardsPerGame","passing_yards_after_catchPerGame","passing_first_downsPerGame",
  "passing_epa","passing_cpoe","passing_2pt_conversionsPerGame","pacr","pass_40_ydsPerGame",
  "pass_incPerGame","pass_comp_pct","fumblesPerGame", #"two_pts",
  "avg_time_to_throw","avg_completed_air_yards","avg_intended_air_yards",
  "avg_air_yards_differential","aggressiveness","max_completed_air_distance",
  "avg_air_yards_to_sticks","passer_rating", #,"completion_percentage"
  "expected_completion_percentage","completion_percentage_above_expectation",
  "avg_air_distance","max_air_distance")
```

```{r}
#efaVars <- c(
#  "completionsPerGame","passing_yardsPerGame","passing_tdsPerGame",
#  "avg_completed_air_yards","avg_intended_air_yards","avg_air_distance",
#  "passing_cpoe","pass_comp_pct","passer_rating","completion_percentage_above_expectation")

#efaVars <- c(
# "passing_yardsPerGame","passing_tdsPerGame",
# "avg_completed_air_yards","avg_air_distance",
# "pass_comp_pct","passer_rating")
#
#cfaVars <- c(
# "completionsPerGame","attemptsPerGame","passing_yardsPerGame","passing_tdsPerGame",
# "passing_air_yardsPerGame","passing_yards_after_catchPerGame","passing_first_downsPerGame",
# "avg_completed_air_yards","avg_intended_air_yards","aggressiveness","max_completed_air_distance",
# "avg_air_distance","max_air_distance","avg_air_yards_to_sticks","passing_cpoe","pass_comp_pct",
# "passer_rating","completion_percentage_above_expectation"
#)

#faVars <- unique(c(cfaVars, efaVars))

efaVars <- c(
 "completionsPerGame","attemptsPerGame","passing_yardsPerGame","passing_tdsPerGame",
 "passing_air_yardsPerGame","passing_yards_after_catchPerGame","passing_first_downsPerGame",
 "avg_completed_air_yards","avg_intended_air_yards","aggressiveness","max_completed_air_distance",
 "avg_air_distance","max_air_distance","avg_air_yards_to_sticks","passing_cpoe","pass_comp_pct",
 "passer_rating","completion_percentage_above_expectation"
)
```

#### Subset Data {#sec-pcaPrepareDataSubset}

```{r}
dataMerged_subset <- dataMerged %>% 
  filter(!is.na(player_id)) %>% 
  filter(position == "QB")
```

@@

Standardizing variables is not strictly necessary in factor analysis; however, standardization can be helpful to prevent some variables from having considerably larger variances than others.

```{r}
dataForFA <- dataMerged_subset
dataForFA_standardized <- dataForFA

dataForFA_standardized[faVars] <- scale(dataForFA_standardized[faVars])
#dataForFA[faVars] <- scale(dataForFA[faVars])
```

@@

cor_mat <- cor(dataForFA[faVars], use = "pairwise.complete.obs")

cor_tidy <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  filter(var1 != var2) %>% # remove self-correlations
  mutate(pair = pmap_chr(list(var1, var2), ~ paste(sort(c(...)), collapse = "___"))) %>% 
  distinct(pair, .keep_all = TRUE) # remove duplicate pairs (symmetry)

high_corrs <- cor_tidy %>%
  filter(abs(correlation) > 0.85) %>%
  arrange(desc(abs(correlation)))

high_corrs

cov_mat <- cov(dataForFA[, faVars], use = "pairwise.complete.obs")
eigen_cov_mat <- eigen(cov_mat)
print(eigen_cov_mat$values)

# Index of smallest (most problematic) eigenvalue
idx <- which.min(eigen_cov_mat$values)

# Extract the associated eigenvector
problem_vector <- eigen_cov_mat$vectors[, idx]
names(problem_vector) <- faVars

# Sort by absolute magnitude to identify most influential variables
sort(abs(problem_vector), decreasing = TRUE)
```

@@

#screeDataEFA <- nFactors::nScree(
#  x = cor( # throws error with correlation matrix, so use covariance instead (below)
#    dataForFA[faVars],
#    use = "pairwise.complete.obs"),
#  model = "factors")

screeDataEFA <- nFactors::nScree(
  x = cov(
    dataForFA[faVars],
    use = "pairwise.complete.obs"),
  cor = FALSE,
  model = "factors")

@@

```{r}
#| label: efa

efa_fit <- lavaan::efa(
  data = dataForFA,
  ov.names = efaVars,
  nfactors = 1:7,
  rotation = "geomin",
  missing = "ML",
  estimator = "MLR",
  bounds = "standard",
  meanstructure = TRUE,
  em.h1.iter.max = 2000000)
```

```{r}
#| label: efa-summary

summary(efa_fit)
```

A path diagram of the EFA model was created using the `lavaanPlot()` function of the `lavaanPlot` package [@R-lavaanPlot].

```{r}
lavaanPlot::lavaanPlot(
  efa_fit$nf3,
  coefs = TRUE,
  stand = TRUE)
```

Here is the syntax for estimating a three-factor EFA using exploratory structural equation modeling (ESEM).
Estimating the model in a ESEM framework allows us to make modifications to the model, such as adding correlated residuals, and adding predictors or outcomes of the latent factors.

```{r}
#| label: efa-3factor-syntax

efa3factor_syntax <- '
 # EFA Factor Loadings
 efa("efa1")*f1 + 
 efa("efa1")*f2 + 
 efa("efa1")*f3 =~ completionsPerGame + attemptsPerGame + passing_yardsPerGame + passing_tdsPerGame + 
 passing_air_yardsPerGame + passing_yards_after_catchPerGame + passing_first_downsPerGame + 
 avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + 
 avg_air_distance + max_air_distance + avg_air_yards_to_sticks + passing_cpoe + pass_comp_pct + 
 passer_rating + completion_percentage_above_expectation
'
```

To fit the ESEM model, we use the `sem()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].

```{r}
#| label: efa-3factor-fit

efa3factor_fit <- sem(
  efa3factor_syntax,
  data = dataForFA,
  information = "observed",
  missing = "ML",
  estimator = "MLR",
  rotation = "geomin",
  bounds = "standard",
  meanstructure = TRUE,
  em.h1.iter.max = 2000000)
```

```{r}
summary(
  efa3factor_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

```{r}
lavaan::fitMeasures(
  efa3factor_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

```{r}
lavaan::residuals(
  efa3factor_fit,
  type = "cor")
```

```{r}
lavaan::modificationindices(
  efa3factor_fit,
  sort. = TRUE)
```

```{r}
lavaan::lavPredict(efa3factor_fit)
```

```{r}
lavaanPlot::lavaanPlot(
  efa3factor_fit,
  coefs = TRUE,
  stand = TRUE)
```

To make the plot interactive for editing, you can use the `plot_lavaan()` function of the `lavaangui` package [@R-lavaangui; @Karch2025_packages]:

@@

lavaangui::plot_lavaan(efa3factor_fit)
```

Here is a modification of the 3-factor model with correlated residuals:

```{r}
#| label: efa-3factorModified-syntax

efa3factorModified_syntax <- '
 # EFA Factor Loadings
 efa("efa1")*f1 + 
 efa("efa1")*f2 + 
 efa("efa1")*f3 =~ completionsPerGame + attemptsPerGame + passing_yardsPerGame + passing_tdsPerGame + 
 passing_air_yardsPerGame + passing_yards_after_catchPerGame + passing_first_downsPerGame + 
 avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + 
 avg_air_distance + max_air_distance + avg_air_yards_to_sticks + passing_cpoe + pass_comp_pct + 
 passer_rating + completion_percentage_above_expectation

 # Correlated Residuals
 completionsPerGame ~~ attemptsPerGame
 passing_yardsPerGame ~~ passing_yards_after_catchPerGame
 attemptsPerGame ~~ passing_yardsPerGame
 attemptsPerGame ~~ passing_air_yardsPerGame
 passing_air_yardsPerGame ~~ passing_yards_after_catchPerGame
 passing_yards_after_catchPerGame ~~ avg_completed_air_yards
 passing_tdsPerGame ~~ passer_rating
 completionsPerGame ~~ passing_air_yardsPerGame
 max_completed_air_distance ~~ max_air_distance
 aggressiveness ~~ completion_percentage_above_expectation
'
```

```{r}
#| label: efa-3factorModified-fit

efa3factorModified_fit <- sem(
  efa3factorModified_syntax,
  data = dataForFA,
  information = "observed",
  missing = "ML",
  estimator = "MLR",
  rotation = "geomin",
  bounds = "standard",
  meanstructure = TRUE,
  em.h1.iter.max = 2000000)
```

```{r}
summary(
  efa3factorModified_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

```{r}
lavaan::fitMeasures(
  efa3factorModified_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

```{r}
lavaan::residuals(
  efa3factorModified_fit,
  type = "cor")
```

```{r}
lavaan::modificationindices(
  efa3factorModified_fit,
  sort. = TRUE)
```

```{r}
lavaan::lavPredict(efa3factorModified_fit)
```

```{r}
lavaanPlot::lavaanPlot(
  efa3factorModified_fit,
  coefs = TRUE,
  stand = TRUE)
```

@@

lavaangui::plot_lavaan(efa3factorModified_fit)
```

```{r}
anova(
  efa3factorModified_fit,
  efa3factor_fit
)
```

@@

psych::principal(
  dataForFA_standardized[faVars],
  nfactors = 3,
  rotate = "oblimin")

'
 #Factor loadings
 F1 =~ completionsPerGame + attemptsPerGame + passing_yardsPerGame + passing_tdsPerGame + passing_interceptionsPerGame + passing_air_yardsPerGame + passing_yards_after_catchPerGame + passing_first_downsPerGame
 F2 =~ pacr + avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + avg_air_yards_to_sticks + expected_completion_percentage + avg_air_distance + max_air_distance
 F3 =~ passing_epa + passing_cpoe + pass_comp_pct + passer_rating + completion_percentage_above_expectation
'

factor1_syntax <- '
 # Factor loadings
 F1 =~ completionsPerGame + attemptsPerGame + passing_yardsPerGame + passing_tdsPerGame + passing_air_yardsPerGame + passing_yards_after_catchPerGame + passing_first_downsPerGame

 # Correlated residuals
 passing_air_yardsPerGame ~~ passing_yards_after_catchPerGame
 completionsPerGame ~~ attemptsPerGame
 attemptsPerGame ~~ passing_yards_after_catchPerGame
 passing_yards_after_catchPerGame ~~ passing_first_downsPerGame
 completionsPerGame ~~ passing_first_downsPerGame
 attemptsPerGame ~~ passing_tdsPerGame
 completionsPerGame ~~ passing_tdsPerGame
 passing_tdsPerGame ~~ passing_air_yardsPerGame
'

factor1_fit <- lavaan::cfa(
  factor1_syntax,
  data = dataForFA,
  missing = "ML",
  estimator = "MLR",
  bounds = "standard",
  std.lv = TRUE,
  cluster = "player_id", # account for nested data within the same player (i.e., longitudinal data)
  em.h1.iter.max = 2000000)

summary(
  factor1_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)

lavaan::modificationindices(
  factor1_fit,
  sort. = TRUE)

factor2_syntax <- '
 # Factor loadings
 F2 =~ avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + avg_air_distance + max_air_distance + avg_air_yards_to_sticks #expected_completion_percentage + pacr

 # Correlated residuals
 max_completed_air_distance ~~ max_air_distance
 avg_completed_air_yards ~~ max_completed_air_distance
 avg_completed_air_yards ~~ avg_air_distance
 avg_completed_air_yards ~~ max_air_distance
 avg_intended_air_yards ~~ max_completed_air_distance
'

factor2_fit <- lavaan::cfa(
  factor2_syntax,
  data = dataForFA,
  missing = "ML",
  estimator = "MLR",
  bounds = "standard",
  std.lv = TRUE,
  cluster = "player_id", # account for nested data within the same player (i.e., longitudinal data)
  em.h1.iter.max = 2000000)

summary(
  factor2_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)

lavaan::modificationindices(
  factor2_fit,
  sort. = TRUE)

factor3_syntax <- '
 #Factor loadings
 F3 =~ passing_cpoe + pass_comp_pct + passer_rating + completion_percentage_above_expectation
'

factor3_fit <- lavaan::cfa(
  factor3_syntax,
  data = dataForFA,
  missing = "ML",
  estimator = "MLR",
  bounds = "standard",
  std.lv = TRUE,
  cluster = "player_id", # account for nested data within the same player (i.e., longitudinal data)
  em.h1.iter.max = 2000000)

summary(
  factor3_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)

lavaan::modificationindices(
  factor3_fit,
  sort. = TRUE)
```

@@

```{r}
cfa1factor_syntax <- '
 #Factor loadings
 F1 =~ completionsPerGame + attemptsPerGame + passing_yardsPerGame + passing_tdsPerGame + 
 passing_air_yardsPerGame + passing_yards_after_catchPerGame + passing_first_downsPerGame + 
 avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + 
 avg_air_distance + max_air_distance + avg_air_yards_to_sticks + passing_cpoe + pass_comp_pct + 
 passer_rating + completion_percentage_above_expectation
'
```

```{r}
#| label: cfa-1factor-fit

cfa1factor_fit <- lavaan::cfa(
  cfa1factor_syntax,
  data = dataForFA,
  missing = "ML",
  estimator = "MLR",
  bounds = "standard",
  std.lv = TRUE,
  cluster = "player_id", # account for nested data within the same player (i.e., longitudinal data)
  em.h1.iter.max = 2000000)
```

```{r}
summary(
  cfa1factor_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

```{r}
lavaan::fitMeasures(
  cfa1factor_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

```{r}
lavaan::residuals(
  cfa1factor_fit,
  type = "cor")
```

```{r}
lavaan::modificationindices(
  cfa1factor_fit,
  sort. = TRUE)
```

```{r}
lavaan::lavPredict(cfa1factor_fit)
```

```{r}
lavaanPlot::lavaanPlot(
  cfa1factor_fit,
  coefs = TRUE,
  stand = TRUE)
```

@@

lavaangui::plot_lavaan(cfa1factor_fit)
```

```{r}
cfa2factor_syntax <- '
 #Factor loadings
 F1 =~ completionsPerGame + attemptsPerGame + passing_yardsPerGame + passing_tdsPerGame + 
 passing_air_yardsPerGame + passing_yards_after_catchPerGame + passing_first_downsPerGame

 F2 =~ avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + 
 avg_air_distance + max_air_distance + avg_air_yards_to_sticks + passing_cpoe + pass_comp_pct + 
 passer_rating + completion_percentage_above_expectation
'
```

```{r}
#| label: cfa-2factor-fit

cfa2factor_fit <- lavaan::cfa(
  cfa2factor_syntax,
  data = dataForFA,
  missing = "ML",
  estimator = "MLR",
  bounds = "standard",
  std.lv = TRUE,
  cluster = "player_id", # account for nested data within the same player (i.e., longitudinal data)
  em.h1.iter.max = 2000000)
```

```{r}
summary(
  cfa2factor_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

```{r}
lavaan::fitMeasures(
  cfa2factor_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

```{r}
lavaan::residuals(
  cfa2factor_fit,
  type = "cor")
```

```{r}
lavaan::modificationindices(
  cfa2factor_fit,
  sort. = TRUE)
```

```{r}
lavaan::lavPredict(cfa2factor_fit)
```

```{r}
lavaanPlot::lavaanPlot(
  cfa2factor_fit,
  coefs = TRUE,
  stand = TRUE)
```

@@

lavaangui::plot_lavaan(cfa2factor_fit)
```

```{r}
anova(
  cfa2factor_fit,
  cfa1factor_fit
)
```

```{r}
cfa3factor_syntax <- '
 #Factor loadings
 F1 =~ completionsPerGame + attemptsPerGame + passing_yardsPerGame + passing_tdsPerGame + 
 passing_air_yardsPerGame + passing_yards_after_catchPerGame + passing_first_downsPerGame

 F2 =~ avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + 
 avg_air_distance + max_air_distance + avg_air_yards_to_sticks

 F3 =~ passing_cpoe + pass_comp_pct + passer_rating + completion_percentage_above_expectation
'
```

```{r}
#| label: cfa-3factor-fit

cfa3factor_fit <- lavaan::cfa(
  cfa3factor_syntax,
  data = dataForFA,
  missing = "ML",
  estimator = "MLR",
  bounds = "standard",
  std.lv = TRUE,
  cluster = "player_id", # account for nested data within the same player (i.e., longitudinal data)
  em.h1.iter.max = 2000000)
```

```{r}
summary(
  cfa3factor_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

```{r}
lavaan::fitMeasures(
  cfa3factor_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

```{r}
lavaan::residuals(
  cfa3factor_fit,
  type = "cor")
```

```{r}
lavaan::modificationindices(
  cfa3factor_fit,
  sort. = TRUE)
```

```{r}
lavaan::lavPredict(cfa3factor_fit)
```

```{r}
lavaanPlot::lavaanPlot(
  cfa3factor_fit,
  coefs = TRUE,
  stand = TRUE)
```

@@

lavaangui::plot_lavaan(cfa3factor_fit)
```

```{r}
anova(
  cfa3factor_fit,
  cfa2factor_fit
)
```

```{r}
cfa3factorModified_syntax <- '
 # Factor loadings
 F1 =~ NA*completionsPerGame + attemptsPerGame + passing_yardsPerGame + passing_tdsPerGame + 
 passing_air_yardsPerGame + passing_yards_after_catchPerGame + passing_first_downsPerGame

 F2 =~ NA*avg_completed_air_yards + avg_intended_air_yards + aggressiveness + max_completed_air_distance + 
 avg_air_distance + max_air_distance + avg_air_yards_to_sticks

 F3 =~ NA*passing_cpoe + pass_comp_pct + passer_rating + completion_percentage_above_expectation

 # Cross loadings
 #F3 =~ attemptsPerGame

 # Correlated residuals
 passing_air_yardsPerGame ~~ passing_yards_after_catchPerGame
 completionsPerGame ~~ attemptsPerGame
 attemptsPerGame ~~ passing_yards_after_catchPerGame
 passing_yards_after_catchPerGame ~~ passing_first_downsPerGame
 completionsPerGame ~~ passing_first_downsPerGame
 attemptsPerGame ~~ passing_tdsPerGame
 completionsPerGame ~~ passing_tdsPerGame
 passing_tdsPerGame ~~ passing_air_yardsPerGame

 max_completed_air_distance ~~ max_air_distance
 avg_completed_air_yards ~~ max_completed_air_distance
 avg_completed_air_yards ~~ avg_air_distance
 avg_completed_air_yards ~~ max_air_distance
 avg_intended_air_yards ~~ max_completed_air_distance

 completionsPerGame ~~ pass_comp_pct
 passing_yardsPerGame ~~ avg_completed_air_yards
 passing_yardsPerGame ~~ max_completed_air_distance
 passing_tdsPerGame ~~ passer_rating
 aggressiveness ~~ completion_percentage_above_expectation

 # Variances
 F1 ~~ 1*F1
 F2 ~~ 1*F2
 #F3 ~~ 1*F3
'
```

```{r}
#| label: cfa-3factorModified-fit

cfa3factorModified_fit <- lavaan::cfa(
  cfa3factorModified_syntax,
  data = dataForFA,
  missing = "ML",
  estimator = "MLR",
  bounds = "standard",
  #std.lv = TRUE,
  cluster = "player_id", # account for nested data within the same player (i.e., longitudinal data)
  em.h1.iter.max = 2000000)
```

```{r}
summary(
  cfa3factorModified_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

```{r}
lavaan::fitMeasures(
  cfa3factorModified_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

```{r}
lavaan::residuals(
  cfa3factorModified_fit,
  type = "cor")
```

```{r}
lavaan::modificationindices(
  cfa3factorModified_fit,
  sort. = TRUE)
```

```{r}
lavaan::lavPredict(cfa3factorModified_fit)
```

```{r}
lavaanPlot::lavaanPlot(
  cfa3factorModified_fit,
  coefs = TRUE,
  stand = TRUE)
```

@@

lavaangui::plot_lavaan(cfa3factorModified_fit)
```

```{r}
anova(
  cfa3factorModified_fit,
  cfa3factor_fit
)
```

### 2025-07-10 getting-started.qmd
20250710 - factor analysis

   install.packages(c(
     "petersenlab","remotes","knitr","rmarkdown","tidyverse","nflverse",
     "tidymodels","easystats","broom","broom.mixed","psych","downlit","xml2",
     "gsisdecoder","progressr","DescTools","pwr","pwrss","WebPower","XICOR",
     "dagitty","ggdag","ggtext","gghighlight","ggExtra","grid","patchwork",
     "pROC", "lme4","lmerTest","MuMIn","emmeans","pbkrtest","sjstats",
     "AICcmodavg","rstan","brms","tidybayes","bbmle","fitdistrplus","sn",
     "mclust","magrittr","viridis","viridisLite","msir","plotly","webshot2",
     "quantmod","fPortfolio","NMOF","nFactors","xts","zoo","forecast","stringi",
     "parallelly","doParallel","missRanger","ggridges","powerjoin","caret",
     "LongituRF","gpboost","corrplot","mgcv","rms","car","lavaan","lavaangui",
     "mice","miceadds","interactions","robustbase","ordinal","MASS",
     "data.table","future","future.apply","SimDesign","domir"))
   ```
1. Some necessary packages, including the [`ffanalytics` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics], are hosted in GitHub (and are not hosted on the Comprehensive R Archive Network [CRAN]) and thus need to be installed using the following code (after installing the `remotes` package [@R-remotes] above)[^petersenlabPackageGitHub]:
   ```{r}
   #| eval: false

>>>>

   install.packages(c(
     "petersenlab","remotes","knitr","rmarkdown","tidyverse","nflverse",
     "tidymodels","easystats","broom","broom.mixed","psych","downlit","xml2",
     "gsisdecoder","progressr","DescTools","pwr","pwrss","WebPower","XICOR",
     "dagitty","ggdag","ggtext","gghighlight","ggExtra","grid","patchwork",
     "pROC", "lme4","lmerTest","MuMIn","emmeans","pbkrtest","sjstats",
     "AICcmodavg","rstan","brms","tidybayes","bbmle","fitdistrplus","sn",
     "mclust","magrittr","viridis","viridisLite","msir","plotly","webshot2",
     "quantmod","fPortfolio","NMOF","nFactors","xts","zoo","forecast","stringi",
     "parallelly","doParallel","missRanger","ggridges","powerjoin","caret",
     "LongituRF","gpboost","corrplot","mgcv","rms","car","lavaan","lavaanPlot",
     "lavaangui","mice","miceadds","interactions","robustbase","ordinal","MASS",
     "data.table","future","future.apply","SimDesign","domir"))
   ```
1. Some necessary packages, including the [`ffanalytics` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics], are hosted in GitHub (and are not hosted on the Comprehensive R Archive Network [CRAN]) and thus need to be installed using the following code (after installing the `remotes` package [@R-remotes] above)[^petersenlabPackageGitHub]:
   ```{r}
   #| eval: false
